---
title: "hw05-ntjjmak"
author: "Nicole Mak"
date: "15/10/2018"
output:
    github_document:
      html_notebook
---

# Assignment #5


**Load the packages to be used and load the singer data set.**

```{r}

suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(gapminder))
```

## Part 1: Factor management

**Factors are how categorical data are stored. The values a factor can take on are called levels. It is important to check variable types. What you think are characters may actually be stored numerically. Let's explore the data frame.**

```{r}
summary(gapminder)
```

**Before beginning the factor management exercise, let's get to know the factors in the gapminder data set.**

```{r}
class(gapminder$country)
class(gapminder$continent)
class(gapminder$year)
class(gapminder$lifeExp)
class(gapminder$pop)
class(gapminder$gdpPercap)

```

**We see that we have two factors in this data set. Let's see the levels for "country" and "continent" before we begin filtering the data.**

```{r}
nlevels(gapminder$country)
nlevels(gapminder$continent)

levels(gapminder$country)
levels(gapminder$continent)
```


**Now let's work towards dropping "Oceoania".**

**We will start by filtering the gapminder data to remove observations associated with the continent of Oceania.**

```{r}
No_Oceania <- gapminder %>%
  filter(continent == "Africa" | continent =="Americas"|continent == "Asia"|continent == "Europe")
No_Oceania %>% 
  sample_frac(0.1) %>% #just showing a sample of the data
  knitr:: kable(format = "markdown", justify = "centre")
```

**Let's check how many levels there are and whether we need to remove unused factor levels.**
```{r}
nlevels(No_Oceania$continent)
levels(No_Oceania$continent)
```

**We see that there are still 5 levels. Let's remove the unused factor levels.**

```{r}
Drop_Oceania <- No_Oceania %>% 
  droplevels()
nlevels(Drop_Oceania$continent)
levels(Drop_Oceania$continent)
```

**We do note that the factor levels are ordered alphabetically. Let's use forcats to re-order the factor levels. We can re-order in different ways.**

**One way to re-order is by frequency.For example, the frequency of continents in the data.**

```{r}
Drop_Oceania$continent %>% 
  fct_infreq() %>% 
  levels()

```

**We can also re-order by the value of other variables in the data as life expectanc or gdp.**

```{r}
Drop_Oceania_1980 <- Drop_Oceania %>% 
  filter(year > 1979) %>% 
  group_by(continent, year) %>% 
  mutate(mediangdp = median(gdpPercap))

Drop_Oceania_1980


fct_reorder(Drop_Oceania_1980$continent, Drop_Oceania_1980$gdpPercap, min) %>% 
  levels() %>% head()
```

**Now we can put this into graph form.**

```{r}
ggplot(Drop_Oceania_1980, aes(x = year, y = Drop_Oceania_1980$mediangdp, color = fct_reorder2(continent, year, Drop_Oceania_1980$mediangdp))) +
  geom_line() +
  labs(color = "Continent")+
  xlab("Year") + ylab("Median GDP") +
  ggtitle("Median GDP 1980-2007 by Continent")
```

**We see that reordering the levels allows the legend to be organized in the same fashion as the trendlines.**


## Part 2 File input and output

**Let's try saving some wrangled data into a new file.**

```{r}
trial<- gapminder %>%
  filter(year == "2007")

write.csv(trial, file = "assignment5_stat545")
```

**Now let's import this file to see if we can read it.**

```{r}
read.csv("assignment5_stat545")
```



## Part 3 Visualization design


**Here is an old graph from assignment 3. It looks at the weighted mean life expectancy for continents over time for the gapminder data set.**

```{r}
gapminder %>% 
 group_by(continent, year) %>%
 summarise(mean_lifeExp_weighted = weighted.mean(lifeExp, pop)) %>% 
 ggplot(aes(year, mean_lifeExp_weighted))+
  geom_point(aes(colour = continent))
```

**What can be improved? 1) We can add a trend line so we can estimate what is happening between years. 2) We can change the axis titles and add a graph title. 3) We can re-order the factor levels so the legend is presented in logical order. 4) We can change the scale of the axes to show every 5 years of life. 5) We can change the graphs theme.**

```{r}
gapminder %>% 
 group_by(continent, year) %>%
 summarise(mean_lifeExp_weighted = weighted.mean(lifeExp, pop)) %>% 
 ggplot(aes(year, mean_lifeExp_weighted, color = fct_reorder2(continent, year, mean_lifeExp_weighted)))+
  geom_point(aes(colour = continent))+
  geom_line(aes(colour = continent))+
  scale_y_continuous(breaks=5*(1:17))+
  xlab("Year")+
    ylab("Weighted Mean Life Expectancy")+
  labs(color = "Continent")+
  ggtitle("Weighted Mean Life Expectancy vs. Time (Years)")+
  theme_minimal()
  
```


**Let's try a different graph and also use `plotly` to make it more interactive.*Now we can make an interactive plot using `plotly`.**

**Here is the original graph:**

```{r}
gapminder %>%
 group_by(year) %>% 
 mutate(median = median(lifeExp)) %>%
 ggplot(aes(year, lifeExp)) +
 geom_jitter(aes(colour = (lifeExp < median)), alpha = 0.5)+
  facet_wrap(~ continent)
```

**Here is the cleaner version.**

```{r}
gapminder %>%
 group_by(year) %>% 
 mutate(median = median(lifeExp)) %>%
 ggplot(aes(year, lifeExp)) +
 geom_jitter(aes(colour = (lifeExp < median)), alpha = 0.5)+
  facet_wrap(~ continent,nrow = 5, ncol = 1)+
  scale_y_continuous(breaks=10*(1:9))+ #change the scale
  xlab("Year")+ #change the titles
  ylab("Life Expectancy")+
  ggtitle("Country life expectancy over time")+
  scale_colour_discrete(name  ="CountrylLife expectancy > worldwide median?")+ #make the legend clearer
  theme(legend.position="right") #select legend position
```


**Now, let's make this into a `plotly` graph.**

```{r}
#First, let's save the above graph into "plotly1"
plotly1 <- gapminder %>%
 group_by(year) %>% 
 mutate(median = median(lifeExp)) %>%
 ggplot(aes(year, lifeExp)) +
 geom_jitter(aes(colour = (lifeExp < median)), alpha = 0.5)+
  facet_wrap(~ continent,nrow = 5, ncol = 1)+
  scale_y_continuous(breaks=10*(1:9))+ 
  xlab("Year")+
  ylab("Life Expectancy")+
  ggtitle("Country life expectancy over time")+
  scale_colour_discrete(name  ="CountrylLife expectancy > worldwide median?")+
  theme(legend.position="right") 

#Now open plotly package.
library(plotly)

#GEnerating plotly graph.
ggplotly(plotly1)

```

**We do notice that `plotly` has some nice features. However, visualization is not as good. The legend is cut off. We can see if moving the legend elsewhere will improve this visualization.

```{r}
plotly2 <- gapminder %>%
 group_by(year) %>% 
 mutate(median = median(lifeExp)) %>%
 ggplot(aes(year, lifeExp)) +
 geom_jitter(aes(colour = (lifeExp < median)), alpha = 0.5)+
  facet_wrap(~ continent,nrow = 5, ncol = 1)+
  scale_y_continuous(breaks=10*(1:9))+ 
  xlab("Year")+
  ylab("Life Expectancy")+
  ggtitle("Country life expectancy over time")+
  scale_colour_discrete(name  ="CountrylLife expectancy > worldwide median?")+
  theme(legend.position="bottom") 

#GEnerating plotly graph.
ggplotly(plotly2)
```

**Hmmmm. It appears `plotly` dose not have all the aesthetic options that ggplot does. Anyways, let's save this as a local html file.**
```{r}
ggplotly(plotly1) %>% 
  htmlwidgets::saveWidget("plotly1")
```

**